\documentclass[]{scrartcl}

\usepackage{amsmath}
\usepackage{bussproofs}
\usepackage{color}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{url}

\EnableBpAbbreviations

\newcommand{\WHILE}[2]{\ensuremath{\mathbf{while}\;#1\;\mathbf{do}\;\{#2\}}}
\newcommand{\IWHILE}[3]{\ensuremath{\mathbf{inv}\;#1\;\mathbf{while}\;#2\;\mathbf{do}\;\{#3\}}}
\newcommand{\IF}[3]{\ensuremath{\mathbf{if}\;#1\;\mathbf{then}\;\{#2\}\;\mathbf{else}\;\{#3\}}}
\newcommand{\SKIP}{\ensuremath{\mathbf{skip};}}
\newcommand{\ASSUME}[1]{\ensuremath{\mathbf{assume}\;#1;}}
\newcommand{\ASSERT}[1]{\ensuremath{\mathbf{assert}\;#1;}}
\newcommand{\WLP}[2]{\ensuremath{\mathbf{wlp}\;#1\;#2}}
\newcommand{\REPBY}[2]{\ensuremath{#1\;\mathbf{repby}\;#2}}
\newcommand{\UNROLLSQ}[3]{\ensuremath{[\mathit{while}]^{#1}(#2, #3)}}
\newcommand{\UNROLLDI}[3]{\ensuremath{\langle\mathit{while}\rangle^{#1}(#2, #3)}}
\lstset{
  frame=none,
  xleftmargin=2pt,
  stepnumber=1,
  numbers=left,
  numbersep=5pt,
  numberstyle=\ttfamily\tiny\color[gray]{0.3},
  belowcaptionskip=\bigskipamount,
  captionpos=b,
  escapeinside={*'}{'*},
  language=haskell,
  tabsize=2,
  emphstyle={\bf},
  commentstyle=\it,
  stringstyle=\mdseries\rmfamily,
  showspaces=false,
  keywordstyle=\bfseries\rmfamily,
  columns=flexible,
  basicstyle=\small\sffamily,
  showstringspaces=false,
  morecomment=[l]\%,
}

%opening
\title{A Program Verification Engine for Imperative Languages}
\subtitle{Course: Program Verification}
\author{Giovanni Garufi, Fabian Thorand}
\date{\today}

\newcommand{\blue}[1]{\textcolor{blue}{#1}}

\newcommand{\HT}[3]{\blue{\{#1\}} #2 \blue{\{#3\}}}

\begin{document}

\maketitle

\tableofcontents

%==================================================================================================
\section{Introduction}
%==================================================================================================

In this report we present our tool for formally verifying programs written in
imperative languages. While testing already significantly decreases the chances
of having bugs in a program, it is often hard or event impossible to write test
cases that cover every execution path.

One way of making sure a program works as intended is to formally prove that its
behavior adheres to the specification.
Such a specification usually consists of a pre- and a post-condition.
The intention is, that the results of a program satisfy the post-condition,
given that the inputs to the program fulfil the pre-condition.
The Hoare calculus provides one mathematical foundation for conducting such
proofs based on pre- and post-conditions.
But of course, doing proofs by hand is error-prone and tedious even for
simple programs.

Ideally, there would be a tool automatically proving properties about programs.
Unfortunately, it is -- in general -- undecidable to check whether a given
program (represented as a Turing machine) satisfies a given property, due to the
Halting problem\footnote{as shown by Rice's theorem}.
The power of turing machines is usually introduced in programming languages
by some form of loops or general recursion.

Fortunately, this can be migitated by requiring the programmer to annotate loops
with invariants, guiding the verification tool in the right direction.
For some specific cases it might even be possible to automatically infer those
invariants.

Using a technique called \emph{Predicate Transformers} it is possible to infer
the precondition required for a program to satisfy its post-condition.

Upon that technique we built our verification engine that we describe in the
following sections.

%==================================================================================================
\section{Project Structure}
%==================================================================================================

We implemented our project in the Haskell programming language making use of
Monads to achieve abstraction.
For invoking the solver we rely on the \texttt{sbv}\footnote{\url{https://hackage.haskell.org/package/sbv}}
library available on Hackage.
It provides a common interface to several solver backends via the SMT-LIB v2 standard,
although some features are only implemented in some solvers.
For that reason, we only use the Microsoft Z3 backend for now.

SBV provides a typed and untyped (w.r.t. expressions) interface to the prover.
As our internal AST representation is untyped and we rely on the type correctness of
the user-provided programs anyway, we use the untyped API.

Our core implementation is devided in four parts, the AST, the monadic DSL, the WLP transformer
and the adapter to the prover backend.
Furthermore, we have implemented a test suite asserting the expected results for the examples
in the project assignment description.

On top of the basic requirements, we also support array assignments, invariant inference and
program calls. The following sections describe our implementation in greater detail.

%==================================================================================================
\section{Monadic DSL for the Guarded Common Language}
%==================================================================================================

Our abstract syntax tree representing programs consists of three levels.
The root node stores the name of the program, the input and output arguments and a body statement.
The statement and expression trees closely follow the grammar outlined in the assignment description,
with one notable difference. Instead of sequencing statements pairwisely, we have an explicit \emph{block}
statement wrapping a list of statements.
Our expression AST is untyped, therefore we require the user to only construct type correct programs.
Variables in the AST are qualified, besides the literal name they also consist of a unqiue identifier.
At certain points in our code we assume that all variables in the supplied AST are indeed unique.

In order to facilitate dealing with variables we also provide a monadic DSL for constructing programs.
The DSL makes use of unqualified names, but performs name resolution and fails if undeclared variables
are used. All variable declarations in the resulting AST are unique.

We also make use of some language features to facilitate writing programs in the DSL.
It is possible to use Haskell string literals to be directly as variables by means of the
\emph{OverloadedStrings} GHC extension and to directly use integer literals and common mathematical
operations by making use the \emph{Num} type class.

It also takes care of some implementation details. If-expressions are actually implemented in terms of
non-deterministic branching in GCL. Furthermore, assignments to arrays like
\lstinline|a[i] := e| are translated to the corresponding \lstinline|a := a(i repby e)|.

Listing \ref{fig:prog:example} shows a program sorting a range $[0\ldots N)$ of an array $a$.
Program calls like \lstinline|call "minind" ["a", "i" + 1, "N"] ["m"]| mean $m := \mathit{minind}(a, i+1, N)$,
that is the first list represents the argument expressions and the second list the return variables.

\begin{lstlisting}[caption=Selection sort implemented in our DSL, label=fig:prog:example]
s = program "sort" [ "a" `as` array int, "N" `as` int] ["ret" `as` array int] $ do
  assume $ 0 .<= "N"
  var ["i" `as` int, "m" `as` int, "min" `as` int] $ do
    "i" $= 0
    let iv = ("N" .== 0 \/ "i" .<= "N" - 1)
            /\ forall ("k" `as` int) (0 .<= "k" /\ "k" .< "i" - 1 ==> 
                 "a" ! "k" .<= "a" ! ("k" + 1))
            /\ forall ("k" `as` int) (0 .<= "k" /\ "k" .< "i" ==> 
                 forall ("l" `as` int) ("i" .<= "l" /\ "l" .< "N" ==> "a" ! "k" .<= "a" ! "l"))
    invWhile (Just iv) ("i" .< "N" - 1) $ do
      call "minind" ["a", "i" + 1, "N"] ["m"]
      if_ ("a" ! "m" .< "a" ! "i")
        (call "swap" ["a", "i", "m"] ["a"])
        skip
      "i" $= "i" + 1
    "ret" $= "a"
    assert $ forall ("k" `as` int) (0 .<= "k" /\ "k" .< ("N" - 1) ==> 
               "ret" ! "k" .<= "ret" ! ("k" + 1))
\end{lstlisting}

%==================================================================================================
\section{Implementing Predicate Transformer}
%==================================================================================================

The WLP transformer is implemented in a custom MonadProver monad we defined.
The MonadProver provides two basic actions: prove and trace, that respectively
either prove or provied a counterexample to a supplied predicate, allowing the
\textbf{wlp} to unfold, and allows us trace debug statements during the whole operation.

As we will see in later sections the there are different ways a WLP computation
could be carried on; these differences are abstracted over by a configuration
that must be provided to the transformer allowing different extensions to the
WLP behaviour.

The transformer operates under the partial correctness interpretation, as such
termination is assumed to be guaranteed.

\subsection{Basic Statements}

Dealing with most of the basic statements in the GCL language turns out to
be pretty straightforward. The implementation deals with these basic statements
exactly as presented in the lectures.
As a matter of fact: since we used the same datatype both for expressions in GCL
and the predicates built by the \textbf{wlp}, we managed to reuse the combinators defined
in the DSL to specify the predicates in the WLP. The effect of this on the code
is that it writes out almost exactly as in the equations presented in the lectures.

Listing \ref{lst:wlp:basic} shows two simple cases from our WLP implementation.
As we need to be able to run the prover on the fly, we run the WLP in a monadic context.
In the case for a block of statements we simply monadicly fold the WLP over all statements.
The implementation for assertions also follows closely the formal specification of the
WLP transformer.

\begin{lstlisting}[caption=WLP implementation example, label=lst:wlp:basic]
  go (AST.Block stmts) q = foldrM go q stmts
  go (AST.Assert e) q =  return (e /\ q)
\end{lstlisting}

\subsection{Loops}

In the case the user provides his loops with an annotated invariant the first
thing that must be done is to verify the validity of said invariant.
This can be stated as
\begin{align}
  I \land \neg g \Rightarrow Q \\
  I \land g \Rightarrow \WLP{S}{I}
\end{align}
The SBV solver has to be invoked to verify that these conditions hold, and that
the invariant supplied by the user is in fact a valid invariant.
If the invariant turns out to be invalid, it can still work out if the loop
is never entered. This gives us the following forumula for the \textbf{wlp} of an annotated
loop
\begin{align}
  \WLP{(\IWHILE{I}{g}{S})}{Q} &=
    \begin{cases}
      I, \qquad  \mathrm{if \; the \; provided \; invariant \; is \; valid}\\
      \neg g \land q, \mathrm{otherwise}
    \end{cases}
\end{align}
We will see in a subsequent section how to deal with unannotated loops.

Listing \ref{lst:wlp:loop} show an excerpt from the WLP function implementing the aforementioned case.
The behavior in the case where the invariant is invalid is configurable. We can either try to prove
that the loop is never entered, or we try to infer a correct invariant.
The invariant check itself can be turned off as well, then the invariant is always assumed correct.
That might be useful when the correctness proof is not feasible with the backend prover.

\begin{lstlisting}[caption=WLP loop case, label=lst:wlp:loop]
go (AST.InvWhile (Just iv) cnd s) q
  | not alwaysInferInvariant && checkInvariantAnnotation = do
      preInv <- go s iv
      let preserveInv = prepare $ iv /\ cnd ==> preInv
          postcnd     = prepare $ iv /\ neg cnd ==> q
      trace "trying to prove invariant is preserved"
      preserved <- prove preserveInv
      trace "trying to prove post condition"
      postValid <- prove postcnd
      if preserved && postValid
        then do
          trace "invariant valid: choosing invariant as precondition"
          return iv
        else case invalidInvariantBehavior of
          NeverExecute -> do
            trace "invariant invalid: requiring that loop is never executed"
            return (neg cnd /\ q)
          Infer -> do
            trace "invariant invalid: trying to infer an invariant"
            inferInv (Just iv) cnd s q
  | not alwaysInferInvariant && not checkInvariantAnnotation = do
      trace "assuming user-supplied invariant is correct"
      return iv
\end{lstlisting}


\subsection{Arrays}

To deal with array assignment we introduce the following translation
\begin{align}
  a[i] := e \qquad \equiv \qquad a := a ( \REPBY{i}{e} )
\end{align}
Where the newly introduced expression represents an array which is identical to
\emph{a}, except at position \emph{i} where it has value \emph{e}.

This allows us to define the corrisponding wlp as if it were a normal assignment
\begin{align}
  \WLP{(a[i] := e)}{Q} &= Q[a( \REPBY{i}{e} ) / a]
\end{align}

\subsection{Program Calls}

We can extend the \textbf{wlp} to verify programs which contain calls to other programs.
We can think of a program call like a statement with two associated sets:
the expressions passed to input argument and the output variables.
External programs can be specified in two different ways: by a complete program or
just by a specification of the preconditions and postconditions associated with
the call.
In the latter case we provide a function that builds the program body by asserting
its pre-conditions and assuming the post-conditions.
This allows us to treat both cases uniformly in the \textbf{wlp} function.
The environment is extended with associations between program names and the
tuple of input/output parameters and program body that completely specify a program
When a program call is made the body is simply inlined at call site taking care
of assigning argument expressions to input parameters and output parameters to the result variables.
Once this program fragment is built the \textbf{wlp} can simply proceed on that.

%==================================================================================================
\section{Invariant Inference}
%==================================================================================================

In the case where the user has not annotated the invariant of while-loops,
it might still be possible to infer the invariant using fixpoint-iteration
or finite unrolling.

\subsection{Fixpoint Iteration}

When we have a loop \lstinline|while g do S;|, it is equivalent to
\begin{lstlisting}
if g then { S; while g do S; }
     else skip;
\end{lstlisting}
effectively unrolling the loop once.

If $W$ is \textbf{wlp} of that loop, we have $W = \WLP{(\WHILE{g}{S})}{Q}$ as well as
$W = (g \land \WLP{S}{(\WLP{(\WHILE{g}{S})}{Q})}) \lor (\neg g \land Q)$ according to the unrolling above.
Therefore, we can conclude that $W = (g \land \WLP{S}{W}) \lor (\neg g \land Q)$.

We can find the greatest fixpoint of this equation by iterating, starting from the weakest invariant \emph{true},
following from the \emph{Knaster-Tarski} fixpoint-theorem.
\begin{align}
  \label{eq:fp:base} W_0 &= \mathit{true} \\
  \label{eq:fp:iter} W_{i+1} &= (g \land W_i) \lor (\neg g \land Q)
\end{align}

If a fixpoint exists, there is an $i$ for which $W_{i+1} = W_i$ holds.
We implemented the iteration according to equations \ref{eq:fp:base} and \ref{eq:fp:iter}.
After each iteration, we invoke the backend-prover to check whether $W_{i+1} \iff W_i$ holds.
In that case we have found the fixpoint $W_i$ and return it to the \textbf{wlp} function as
invariant.
We additionally implemented an optional limit to the number of iterations.
If that limit is exceeded without finding an invariant, we return a precondition that requires
that the loop will never be executed and the post-condition already holds.

\subsection{Loop Unrolling}

As an alternative to fixpoint-iteration we can also unroll a loop a finite number of times and
compute the \textbf{wlp} of the unrolling. The last iteration then either asserts that the
loop guard will not hold, or just assume that it doesn't hold anymore.
Formally, those unrollings, named  are defined as follows.
\begin{align}
\UNROLLSQ{0\phantom{+1}}{g}{S} &= \ASSERT{\neg g} \\
\UNROLLSQ{k+1}{g}{S} &= \IF{g}{S; \UNROLLSQ{k}{g}{S}}{\SKIP}
\end{align}

\begin{align}
\UNROLLDI{0\phantom{+1}}{g}{S} &= \ASSUME{\neg g} \\
\UNROLLDI{k+1}{g}{S} &= \IF{g}{S; \UNROLLDI{k}{g}{S}}{\SKIP}
\end{align}

Our implementation first computes (lazily) the unrolled loop and then applies
the \textbf{wlp} transformer to the unrolling.

%==================================================================================================
\section{Solver Backend}
%==================================================================================================

As mentioned before we rely on the \texttt{sbv} Haskell library for interacting with a theorem prover.
While it supports multiple provers, we only tested our implementation with Microsoft's Z3 being the
one with most features.
The WLP implementation itself is not bound to a specific prover backend and it should be easy to
add others if necessary.

\subsection{Free Monad Interface}

The WLP transformer is implemented as a free monad over WLP computations.
This allows us to specify different interpreters for the same abstract WLP tree.
We ended up implementing a handful of different interpreters along the ``main''
one, which relies on the SBV interface to carry out the proofs in predicate logic.
Other interpreters that were implemented allowed us to: \textit{pretty-printer} of WLP trees,
allowing us to inspect the different branching points and in general providing
a very useful debugging tool, an \textit{interactive} interpreter which asks the user
to carry out the requested proofs and a \textit{speculative} interpreter which, at branching
points, immediately proceeds resolving the \textbf{wlp} with all possible outcomes and
at a later point prunes the excessive branches when a proof is provided.

\subsection{Handling Quantifiers}

Our GCL DSL allows arbitrary nesting of universal and existential quantification,
but the solver library we use requires all quantified variables to be defined in
the beginning.
Fortunately, for every formula in classical logic there is an equivalent formula
in prenex normal form, where all quantifiers occur in the beginning.
The conversion to prenex normal form requires that every quantified variable is
unique to avoid accidental capturing when pulling the quantifiers outside.
Additionally, all negations have to be pushed inwards, swapping universal and
existential quantification on the go.
Skolemization is then perfomed by the prover backend.
An additional benefit from the conversion to prenex normal form is, that it allows
us to return a model for all existential quantifications in a SAT proof, as only
the values of the outermost existentials are returned by the library.

\subsection{Array Theory}

While the conversion and meaning of integer arithmetic and boolean logic should be
straightforward, arrays require are more subtle treatment.
As explained above, array assignments are translated to the $a(\REPBY{i}{e})$ syntax.
Fortunately, Z3 provides a built-in array theory with two functions \emph{select} and
\emph{store} directly mapping to array indexing and \emph{repby}.
The downside is, that the \texttt{sbv} library we use imposes some restrictions on the
use of arrays and it was not clear to us from the documentation whether these restrictions
stem directly from Z3 or from the wrapper library.
Most notable, arrays can only be used existentially during SAT proofs (i.e. only universally
during validity proofs).
This didn't mean any restriction for our examples though, because all variable declarations
result in universal quantifications.

\section{Conclusion}

Using Haskell allowed us to write the WLP implementation in a succinct declarative
style, closely following the underlying formal definitions. It also enabled us
to easily define a custom DSL for the GCL language which made working on the GCL
programs a relatively pain-free experience. 

\end{document}
